Urban Connectivity Dynamics: Analyzing Free Public WiFi Distribution in Toronto

### 2024-04-16 outputs/paper/paper.qmd
revise the structure

---
title: "Better Talent and Rule Changes Have Resulted in More Goals in the NHL"
subtitle: "An Analysis of the Increase in NHL Goal Scoring From 2011 to 2023"
author: 
  - Jason Ngo
thanks: "Code and data are available at: https://github.com/thejasonminh/nhlpoints"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
format: 
  pdf:
    documentclass: article
    geometry: margin = 1in
    abstract: "Of the four major sports in the USA, hockey is the least popular. In an effort to grow popularity, the NHL (National Hockey League) strove to increase goal scoring. Using data from the NHL, this paper will analyze average leaguewide goal scoring per game from the 2011-2012 to 2022-2023 seaons. A linear regression model will also be created to predict goal scoring averages in the upcoming season. Results show that the NHL was succesful as the average goals per game increased over a 10 year period. These findings suggest that increased player talent and rule changes are primary factors."
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
library(knitr)

league_data <- read.csv(here::here("outputs/data/league_data.csv"))
master_dmen <- read.csv(here::here("outputs/data/master_dmen.csv"))
master_forwards <- read.csv(here::here("outputs/data/master_forwards.csv"))
goals_model <- read_rds(here::here("outputs/models/goals_model.rds"))
```

# Introduction
Ice Hockey, Baseball, Football, and Basketball have become so popular in the United States, they've become branded as "The Big Four." The top leagues that operate in each sport: The NHL, MLB, NFL, and NBA are all centralized in the USA. Many of the top players that have played, or currently play, in these leagues are American, and their popularity only serves to inspire more youth to follow in their footsteps. There are no other sports that can threaten the stranglehold that these four sports have on the USA.

However, hockey has fallen behind the other three. Viewership, revenue and general popularity all pale in comparison to the other three sports [@citestatista]. As a result; player salaries are lower in comparison to other sports and there is less youth talent in the pipelines. Two of the biggest factors for this decline are: High price of entry and low scoring. @citegear details the high upfront price of equipment and the amount of equipment required to play ice hockey. Because of this high upfront price, many kids elect not to choose hockey as their sport and instead elect for cheaper options such as basketball. Low scoring was a result of goalies having too much freedom, and teams creating bulletproof defensive tactics that involved clogging the middle of the ice and using their goalie to stifle momentum.

As a response to the declining popularity, the NHL decided to implement rule changes and goalie equipment resizing to facilitate the increase of goal scoring. By increasing goal scoring, games would be more exciting to watch which in turn, would lead to increased viewership. This paper will analyze the effectiveness of these changes by drawing data from the NHL to analyze the estimand: Average leaguewide goals scored per game from the 2011-2012 season to the current (2022-2023) season. In addition, the paper will draw on a linear regression model to predict average goal scoring for the future season.

The findings show an increase of average goals scored per game over a 12 season period, with this current season exhibiting the highest goals scored per game during this period. The rule changes have given players more freedom to play and goalie equipment resizing has made goalies more susceptible to getting scored on. In addition, player talent has greatly increased over the years, with both forwards and defensemen exhibiting drastic offensive growth.

In @sec-data of the paper, the source of the data as well as the three primary datasets are discussed. Strengths and weaknesses, methods of collection and terminologies will be explored. @sec-model of the paper will introduce a linear model, its justifications, and how it was constructed. In addition, the model will predict goal scoring averages for next season. @sec-results will discuss the results, touching on the trends and events that occurred over a 12-season period. @sec-discussion of the paper will touch on the factors that contributed to increased goal scoring as well as weaknesses and next steps.

# Data {#sec-data}
There are three primary datasets used for this paper. The first dataset, @sec-leagueavg, contains league averages from NHL seasons 2011-12 to 2022-23. The next dataset, @sec-fwds, contains the average points-per-game for every forward from the 2011-12 to 2022-23 seasons with a recorded average points-per-game pace of 1.00 or above. The last dataset, @sec-dmen, contains the average points-per-game for every defenseman from seasons 2011-12 to 2022-23 with a recorded average points-per-game pace of 0.60 or above.

The analysis of these three datasets will be carried out using the statistical programming language `R` [@citeR], using the `tidyverse`[@citetidy], `here`[@citehere], and `readxl`[@citexl] packages. The figures and tables in the paper are generated using the, respectively, `ggplot2`[@citegg] and `knitr`[@citeknitr] packages.

## League Averages Dataset {#sec-leagueavg}

The league averages dataset, represented below in @tbl-1, is created through averaging each variable across every team in the league for the corresponding season. Seasons 2011-12 to 2016-17 has averages for 30 teams, seasons 2017-18 to 2020-21 has averages for 31 teams, and seasons 2021-22 and 2022-23 has averages for 32 teams. Each team is responsible for tracking their own statistics, which is then conglomerated and represented by both the NHL and third party websites. Statistics tracking is done during real time, while a game is in progress. Internal models are used to calculate percentage, average and advanced statistics.

With every team using their own methods for tracking statistics, measurement bias comes into play. Because the NHL does not standardize methods for tracking statistics, some teams will skew statistics such that their team is represented better.

```{r}
#| label: tbl-1
#| tbl-cap: The dataset containing averages from NHL seasons 2011-2012 to 2022-2023.
#| echo: false
#| warning: false

kable(league_data, padding = 5)
```

Almost all of @tbl-1 is taken from @citehref, with the exception of the `Sh_Pct` column, which was pulled from @citenhl and merged with the larger dataset. The reasoning why I didn't use the NHL website for the majority of this dataset was that the NHL website did not support automatic calculation of league averages. There are 8 columns total in this league averages dataset:

1. The `Season` column, which denotes the NHL season each average represents. 
2. The `Goals_PG` column, which denotes the average goals per game for the respective season. 
3. The `PP_PG` column, which denotes the average power plays, situations where one team has a man-advantage due to a penalty, per game for the respective season. 
4. The `PP_Pct` column, which shows the percentage of power plays that were successful, or resulted in a goal, for the respective season. 
5. The `SA` column, which details the leaguewide average shots a team faced per game, for the respective season.
6. The `SV_Pct` column, which denotes the leaguewide average save percentage, which represents the percentage of total saves made, per season.
7. The `GA_Avg` column, which represents the leaguewide average goals against per season.
8. The `Sh_Pct` columnm which represents the leaguewide average shooting percentage. Shooting percentage represents the percentage of shots that are goals.

The `Season` column was necessary as a timeframe is needed to anchor this analysis. The `Goals_PG` was chosen as it is the estimand of this paper. The remaining 6 columns were chosen as they all correlate to goal scoring, or situations that incentivizes goal scoring. For example, having a high shooting percentage correlates to having a higher goals per game, which would decrease the save percentage and increase the goals against average. The powerplay columns represent the opportunities teams have to score on the man advantage and the capability for teams to score on that advantage.

## Forwards Dataset {#sec-fwds}

The forwards dataset, represented below in @tbl-2, represents the count of every forward from the 2011-12 to 2022-23 seasons that finished the season with, at minimum, a 1 point-per-game (PPG) pace. A point represents either a goal or an assist. The reason why I chose to filter the forwards by the 1 PPG pace is because there are few players that finish the season at a 1 PPG pace, which makes those players that are able to do so offensively gifted.

The calculation of average points-per-game is quite simple: Simply take a player's total points and ratio it against the amount of games they've played. It is very simple to track a player's points since every game log has every goal with their respective assists recorded. This is true for both forwards and defensemen.

```{r}
#| label: tbl-2
#| tbl-cap: The count of forwards that finish at a 1 PPG pace or above per season.
#| echo: false
#| warning: false

kable(master_forwards |> count(Season) |> rename("PPG Players" = "n"), padding = 5)
```

All the data about forwards was taken from [@citenhl]. Since we are only concerned with the average offensive output of forwards, I chose to only analyze the points-per-game pace of forwards.

## Defensemen Dataset {#sec-dmen}

The defensemen dataset, represented below in @tbl-3, represents the count of every defensemen from the 2011-12 to 2022-23 seasons that finished the season with, at minimum, a 0.6 point-per-game pace. Rather than filtering by the 1 PPG pace like I did for forwards, I chose to filter defensemen by a 0.6 PPG pace. The reasoning for doing so is because defensemen simply do not contribute offensively as much as forwards do. A defensemen will rarely match a forward's offensive output because offense isn't the primary objective of a defensemen. However, talented defensemen that are table to contribute offensively usually do so at a 0.6 point-per-game pace or above.

```{r}
#| label: tbl-3
#| tbl-cap: The count of defensemen that finish at a 0.6 PPG pace or above per season.
#| echo: false
#| warning: false

kable(master_dmen |> count(Season) |> rename("PPG Players" = "n"), padding = 5)
```

All the data regarding defensemen was also taken from [@citenhl]. Similar to the forwards, I am only concerned with the offensive output of defensemen so I chose to only analyze the point-per-game pace of defensemen.

# Model {#sec-model}

The goal of my modelling strategy is twofold. Firstly, I want to analyze to what effect power play opportunities, power play effectiveness, save percentage, goals against average and shooting percentages all have on average goals scored per game. Secondly, I want to use these results to predict the average goals scored per game for the next season.

In the section below, the Bayesian analysis model used to investigate the multiple linear regression model of average goals scored per game is detailed.

## Model set-up

Define $y_i$ as the leaguewide average goals scored per game. Then $\beta_1$ is the leaguewide average power plays per game, $\gamma_1$ is leaguewide powerplay percentage, $\theta_1$ is leaguewide save percentage, $\phi_1$ is the leaguewide goals against average, and $\kappa_1$ is the leaguewide average shooting percentage.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \theta_i + \phi_i + \kappa_i + \beta_i * \gamma_i \\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\theta &\sim \mbox{Normal}(0, 2.5) \\
\phi &\sim \mbox{Normal}(0, 2.5) \\
\kappa &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in `R` [@citeR] using the `rstanarm` package of [@citestan]. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between average goals scored per game and power play opportunities, power play percentage, goals against average, and shooting percentage. The higher those four variables are, the more goals there should be. In particular, there should also be a negative relationship between goals scored and save percentage. With a lower save percentage, there are less saves that are occuring which means there should be more goals. Additionally, the two power play variables are interacted with eachother, since they are directly related to one another. There cannot be a power play percentage without a power play opportunity.

## Model Prediction
Taking the differences in averages from the 2021-2022 season and the 2022-2023 season, I will create a test dataset for the 2023-2024 season by adding those differences onto the averages from the 2022-2023 season and use the model with this test dataset to predict the average goals scored per game for the 2023-2024 season. @tbl-4 is the resulting mock 2023-2024 season averages data.
```{r}
#| echo: false
#| warning: false

test_league <- tibble(
  Season = c(
    "2023-24"
    ),
  Goals_PG = c(
    3.22
  ),
  PP_PG = c(
    3.25
  ),
  PP_Pct = c(
    22.01
  ),
  SA = c(
    30.8
  ),
  SV_Pct = c(
    0.901
  ),
  GA_Avg = c(
    3.02
  ),
  Sh_Pct = c(
    10.33
  )
)

prediction <- predict(goals_model, newdata = test_league)
```

```{r}
#| label: tbl-4
#| tbl-cap: A mock league averages dataset for the 2023-2024 NHL season.
#| echo: false
#| warning: false

kable(test_league, padding = 5)
```

Using the mock data alongside the model, the predicted average goals per game for the 2023-2024 season is **3.21**.

@@

# Results {#sec-results}

## Goals, Saves and Shooting

```{r}
#| label: fig-1
#| fig-cap: An increase of average goals scored per game over 12 seasons.
#| echo: false
#| warning: false

library(tidyverse)

league_data |> ggplot(aes(x = Season, y = Goals_PG)) +
  geom_line(group = 1) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Leaguewide Average Goals Scored Per Game over 12 Seasons",
    x = "Season",
    y = "Average Goals P/G"
  ) 
```

@fig-1 shows an upward trajectory of goal scoring per season. The horizontal axis denotes the season, and the vertical axis denotes the average goals scored per game. There are two things of note: First, there was a big spike in goal scoring starting in the 2017-18 season, with an increase of 0.2 from the 2016-17 season. with the subsequent seasons building on that jump. The second thing of note however, is the slight dip in scoring during the 2020-21 season where the average fell to 2.94 from the 3.02 the season before.

In regards to the spike of scoring in the 2017-18 season, the NHL had implemented a few rule changes that helped facilitate scoring. Additionally, goalie equipment had shrunken by the time the season started, so goalies were scored on more.

With respect to the slight dip in scoring in the 2020-21 season, that season took place during the height of the COVID-19 pandemic and due to health regulations, many players had to miss extended periods of time to self-quarantine after a positive test. The loss of skilled players for periods of time and the erosion of team chemistry led to the slight decrease in goals scored that season.

```{r}
#| label: fig-2
#| fig-cap: Implications on increased goal scoring on save percentage, goals against average and shooting percentage.
#| fig-subcap: ["Save Percentage", "Goals Against Average", "Shooting Percentage"]
#| echo: false
#| warning: false
#| layout-ncol: 2

library(tidyverse)

league_data |> ggplot(aes(x = Season, y = SV_Pct)) +
  geom_line(group = 1) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Leaguewide Save Percentage over 12 seasons",
    x = "Season",
    y = "Average Save %"
  )

league_data |> ggplot(aes(x = Season, y = GA_Avg)) +
  geom_line(group = 1) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Leaguewide Goals Against Average over 12 seasons",
    x = "Season",
    y = "GAA"
  )

league_data |> ggplot(aes(x = Season, y = Sh_Pct)) +
  geom_line(group = 1) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Leaguewide Average Shooting Percentage over 12 seasons",
    x = "Season",
    y = "Average Sh %"
  ) 
```

@fig-2 shows the implications of increased goal scoring in average save percentage, goals against average and average shooting percentage. A better shooting percentage means more goals are scored. More goals scored increases the goals against average, and more goals scored means less saves are made.

## More Talent, Better Power Plays

```{r}
#| label: fig-3
#| fig-cap: The rise of offensively gifted forwards and defensemen over 12 seasons.
#| fig-subcap: ["Forwards", "Defensemen", "Shots Per Game"]
#| echo: false
#| warning: false
#| layout-ncol: 2

library(tidyverse)

master_forwards |> count(Season) |> ggplot(aes(x = Season, y = n)) +
  geom_line(group = 1) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Productive Forwards over 12 seasons",
    x = "Season",
    y = ">= 1 PPG Forwards"
  )

master_dmen |> count(Season) |> ggplot(aes(x = Season, y = n)) +
  geom_line(group = 1) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Productive Defensemen over 12 seasons",
    x = "Season",
    y = ">= 0.6 PPG Defensemen"
  )

league_data |> ggplot(aes(x = Season, y = SA)) +
  geom_line(group = 1) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Avg. Shots Per Game over 12 seasons",
    x = "Season",
    y = "Shots Per Game"
  ) 
```

@fig-3 showcases the increase of offensively talented forwards and defensemen over 12 seasons, with an additional graph of leaguewide shots per game to contextualize the increase in talent. For offensively talented forwards, the horizontal axis denotes the season while the vertical axis denotes the amount of forwards that finished the season at a pace of 1 PPG or greater. Similarly, the offensively talented defensemen graph has the same horizontal axis, while the vertical axis denotes the amount of defensemen that finished the season at a pace of 0.6 PPG or greater. For shots per game, the horizontal axis once again denotes the season and the vertical axis denotes the average leaguewide shots per game.

Similar to the above section about goal scoring, the shift from the 2016-17 to the 2017-18 season brought an uptick in scoring alongside it. Regarding forwards, it is interesting to note that 2012-13 total of 20 players was never reached again until the 2017-18 season. The 2014-15, 2015-16, and 2016-17 seasons all saw a drop in productive forwards. This trend is not reflected in the defensemen graph. Aside from a slight drop in the 2016-17 season, the amount of offensively productive defensemen slowly increased over the years. This trend means that over the seasons, defensemen have gotten more involved with the offense. This could be a result of increased talent, or a shifting in tactics to allow defensemen to flourish offensively, or perhaps a mixture of both.

However, what makes this increase in productivity even more substantial is looking at the average shots per game over 12 seasons. Over the 12 season period, the lowest average shots per game was 29, and thee highest was 31.8. This means that teams are barely shooting more; the average shots per game over 12 seasons has remained somewhat consistent. Players are simply doing more with their shots on goal.

```{r}
#| label: fig-4
#| fig-cap: The implications of better talent on power play effectiveness.
#| fig-subcap: ["Amount of Power Plays", "Effectiveness of Power Plays"]
#| echo: false
#| warning: false
#| layout-ncol: 2

library(tidyverse)

league_data |> ggplot(aes(x = Season, y = PP_PG)) + 
  geom_line(group = 1) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Power Play Opportunities over 12 seasons",
    x = "Season",
    y = "Avg. PP Opportunities"
  )

league_data |> ggplot(aes(x = Season, y = PP_Pct)) + 
  geom_line(group = 1) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Power Play Effectiveness over 12 seasons",
    x = "Season",
    y = "Avg. PP %"
  ) 
```

One consequence of increased offensive talent is a greater effectiveness of the power play. @fig-4 shows that, despite a decrease in overall power plays over a 12 season period, power play effectiveness has trended upwards. In essence, players are making more with less, and are doing so because of how much more skilled players are in comparison to players before.

# Discussion {#sec-discussion}

As evidenced by the results, goal scoring has steadily increased over the seasons. The NHL was successful in their goal of making the game more exciting, but how did they go about doing it? The two primary factors that led to this increase was an increase of mid-to-high tier player talent, as well as rule changes.

## Good Players Becoming Great

While I talked about the increase of the amount of productive forwards and defensemen, here I will discuss the increase in the amount of good forwards and compare them to the amount of elite forwards. To be classified as an elite forward, usually they would need to finish the season at around a 1.4 PPG pace. Because of how hard of a feat it is to achieve a sustained 1.4 PPG pace, there are very few elite forwards in the NHL. @fig-5 shows the amount of forwards that finish in multiple PPG ranges.

```{r}
#| label: fig-5
#| fig-cap: Different Forward PPG brackets over 12 seasons.
#| fig-subcap: ["1.4 or above", "Between 1.2 and 1.4", "Between 1.0 and 1.2"]
#| echo: false
#| warning: false
#| layout-ncol: 2

library(tidyverse)

master_forwards |> filter(Points_PG >= 1.4) |> count(Season) |> ggplot(aes(x = Season, y = n)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Forwards above 1.4 PPG over 12 seasons",
    x = "Season",
    y = "Player Count"
  )

master_forwards |> filter(Points_PG >= 1.2 & Points_PG < 1.4) |> count(Season) |> ggplot(aes(x = Season, y = n)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Forwards Between 1.2 and 1.4 PPG over 12 seasons",
    x = "Season",
    y = "Player Count"
  )

master_forwards |> filter(Points_PG >= 1.0 & Points_PG < 1.2) |> count(Season) |> ggplot(aes(x = Season, y = n)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Forwards Between 1.0 and 1.2 PPG over 12 seasons",
    x = "Season",
    y = "Player Count"
  ) 
```

As evidenced, the amount of elite players have remained consistent over the seasons, although the big spike in the 2021-22 season may be a sign that times are changing. The bigger story here though, is the amount of players that finish in the 1.2 and 1.4 range from 2017-18 beyond. There is a dramatic spike in these productive players from before 2017-18 to that season and beyond. The graphs here show that there while the ceiling of player talent has remained the same, the floor is steadily increasing.

The same level of growth can be seen with defensemen as well. @fig-6 shows the amount of defensemen that finish in multiple PPG brackets.

```{r}
#| label: fig-6
#| fig-cap: Different Defenseman PPG brackets over 12 seasons.
#| fig-subcap: ["0.95 or above", "Between 0.75 and 0.95", "Between 0.6 and 0.75"]
#| echo: false
#| warning: false
#| layout-ncol: 2

library(tidyverse)

master_dmen |> filter(Points_PG >= 0.95) |> count(Season) |> ggplot(aes(x = Season, y = n)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Defensemen above 0.95 PPG over 12 seasons",
    x = "Season",
    y = "Player Count"
  )

master_dmen |> filter(Points_PG >= 0.75 & Points_PG < 0.95) |> count(Season) |> ggplot(aes(x = Season, y = n)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Defensemen Between 0.75 and 0.95 PPG over 12 seasons",
    x = "Season",
    y = "Player Count"
  )

master_dmen |> filter(Points_PG >= 0.6 & Points_PG < 0.75) |> count(Season) |> ggplot(aes(x = Season, y = n)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  labs(
    title = "Defensemen Between 0.6 and 0.75 PPG over 12 seasons",
    x = "Season",
    y = "Player Count"
  ) 
```
There is a slightly different trajectory with defensemen. It appears that while the floor of offensive production has remained somewhat consistent over the seasons, the ceiling has only gotten higher. It appears that we may be in the middle of a renaissance of offensively gifted defensemen, with multiple defensemen over the last two seasons exhibiting a PPG of 0.95 or above. To contextualize, a great forward should finish at a 1.00 PPG, so to see multiple defensemen reach those numbers is unprecedented.

The floor of talent for forwards have increased, and the ceiling of talent for defensemen have increased, and both of these factors contributed to a spike in goal scoring.

## Changes in 2017-18

While the findings have shown a general increase in goal scoring, the biggest spike in productivity lies between the 2016-17 season and the 2017-18 season. While blossoming player talent is certainly a factor here, the biggest factor in play are rule changes and goalie equipment changes.

### Rule Changes

>>>>

---
title: "Urban Connectivity Dynamics: Analyzing Free Public WiFi Distribution in Toronto"
subtitle: "By Building Typology, Population, and Socioeconomic Indicators"
author: 
  - Bernice Bao
thanks: "Code and data are available at: <https://github.com/BerniceBao/Free-Public-WiFi-Distribution-in-Toronto>
 <https://open.toronto.ca/dataset/city-of-toronto-free-public-wifi/>, and <https://open.toronto.ca/dataset/ward-profiles-25-ward-model/>."
date: 15 April 2024
date-format: long
abstract: "In the era of increasing digital dependence, understanding the intricacies of urban internet connectivity is pivotal for fostering an effective and technologically equitable society. Public WiFi is a vital tool in modern society, offering internet access and supporting economic development, education, and social connections. In response to the increasing importance of internet connectivity in contemporary urban life, I investigate the distribution of free public WiFi in Toronto. This paper aims to provide key statistics on the distribution of public WiFi in Toronto and analyze the impact of building types, demographics, and economic influences on its distribution to gain as comprehensive a picture as possible of the role and impact of public WiFi in our lives."
format: pdf
fig-pos: "H"
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false

#### Workspace set-up ####
library(tidyverse)
library(lubridate)
library(knitr)
library(palmerpenguins)
library(dplyr)
library(here)

#### Read in the data ####
cleaned_wifi_data <-
  read_csv(
    file = here("outputs/data/cleaned_wifi_data.csv"),
    show_col_types = FALSE
  )
#### Read in the data ####
cleaned_ward_data <-
  read_csv(
    file = here("outputs/data/cleaned_ward_data.csv"),
    show_col_types = FALSE
  )
```

# Introduction {#sec-intro}

In an era characterized by an ever-deepening reliance on digital technologies, the intricate tapestry of urban internet connectivity is emerging as a pivotal force in the transformation of our societal fabric. Countries globally are acknowledging internet access as an indispensable tool for active participation in contemporary democratic societies. In Canada, the Canadian Radio-television and Telecommunications Commission (CRTC) has recognized that "broadband internet access services are crucial to Canada’s economic, social, democratic, and cultural fabric."(@citiesfordigitalrights-toronto-wifi) As a university student navigating the complexities of contemporary living, the significance of understanding these dynamics cannot be overstated. We are acutely aware of the transformative potential that connectivity holds.

Public WiFi, often taken for granted in our daily lives, emerges as a vital tool shaping the contours of a technologically inclusive society. It plays a pivotal role in modern society, offering a myriad of benefits to individuals.  Beyond providing a cost-effective alternative to cellular data, it fosters internet accessibility, enabling people to stay connected without constraints.  This flexibility and mobility are particularly advantageous for students, travelers, and those on a budget, allowing them to work, study, or communicate from diverse locations.  Moreover, public WiFi contributes to social inclusion, bridging the digital divide and ensuring that even those without home internet access can engage in online activities.  Access to information, job searching, and remote work opportunities are further facilitated, enhancing the overall quality of life.  In emergencies, public WiFi serves as a crucial communication lifeline, demonstrating its indispensable role in connecting people and providing access to vital services.(@broadbandsearch-public-wifi)

Beyond the conveniences of internet access, it serves as a linchpin fostering economic development, facilitating education, and fortifying social connections. The present study, anchored in the evolving landscape of Toronto, endeavors to unravel the mysteries surrounding the distribution of public WiFi. This paper embarks on a compelling exploration of the distribution of free public WiFi in Toronto, recognizing it as not merely a technological amenity but a cornerstone of our modern existence. By examining the impact of various factors, such as building types, demographic patterns, and economic influences, this research seeks to paint a comprehensive portrait of the role and impact of public WiFi in our lives, providing readers with a lens to comprehend the symbiotic relationship between technology and urban living.

This research is divided into three main parts: `Introduction`(@sec-intro), `Data`(@sec-data), and `Conclusion`(@sec-conclusion). Commencing with the Data section, a meticulous exposition ensues, delineating the origin of datasets garnered from the OpenDataToronto Library and expounding upon the procedural rigors applied for data refinement and analysis. The conclusion part shows what I found during the analysis, while goes deeper into those findings, and finally wraps up the main discoveries from this paper about public free WiFi in Toronto.

# Data {#sec-data}

In this section, I will explore the diverse modes of `R`-based data collection and generation within the `RStudio` environment [@citeR; @citeRStudio]. The data under scrutiny in this paper originates from the `opendatatoronto` R package [@citeODT], specifically drawing from the Toronto fire incidents dataset. The analysis benefits from additional functionalities harnessed from various R packages such as `tidyverse` [@citeTidyverse], `here` [@citeHere], `readr` [@citeReadr], `ggplot2` [@citeGgplot2], `knitr` [@citeKnitr], and `dplyr` [@citeDplyr]. Further insights into the deployment of these packages will be expounded upon in the ensuing subsections.

## Location and description data of Toronto free public WiFi

The data I mainly used to generate the visualizations in this paper is City of Toronto Free Public WiFi from Open Data Toronto [@citeODT] that is an open source tool designed to be publicly available and encourage the development of valuable insights. The dataset, released by Toronto's Department of Information & Technology, captures information on location and description data free public WiFi locations in the City of Toronto and is updated monthly.

The data used for analysis in this article is as of January 3, 2024. According to the data characteristics described on the portal, this data should include information such as address, building type, building name, postal code, whether there is public wifi, etc. Through systematic analysis, it is found that only building type is the main subject variable that can be used to study the influencing factors of distribution. In order to quickly extract available data, only columns with reasonable data are retained during the data cleaning process (see Table 1). Basic data cleaning of column values to shorten the description of different types of clinics and improve readability. I create a statistic on the basis of monthly groups, using kable() from knitr to create @tbl-wifi.

```{r}
#| label: tbl-wifi
#| tbl-cap: Sample of cleaned Free Public WiFi data
#| echo: false

cleaned_wifi_data %>%
  head(10) %>%
  kable(
    col.names = c("Building type", "Ward", "Ward Number"),
    digits = 1
  )

```

From the Location and description data of Toronto free public WiFi, as of January 3, 2024, there are 194 free public WiFi location in the City of Toronto. Further breakdown shows that there are 27 free public WiFi location that are set in Arena, 74 free public WiFi location that are set in Community Recreation Centre, 17 free public WiFi location that are set in Outdoor Pool Building clinics, and  100 free public WiFi location that are set in library. Since the same wifi may cover multiple buildings, the sum of the number of public wifi corresponding to each building type is greater than the total number of public wifi in Toronto. The four building types mentioned above have the largest number of public wifi, covering almost 90% of total.

Library has the highest number of free public WiFi, which means that, in comparison to other types of buildings in Toronto, libraries offer the highest availability or coverage of free public WiFi services. This suggests that libraries prioritize or invest in providing accessible and free internet connectivity to the public within Toronto. Besides, Community Recreation Centre has the second most amount of free public WiFi. This information is indicative of the efforts or infrastructure investments made by Community Recreation Centres to offer accessible and free internet connectivity to the public.

## 2021 Ward Profiles based on the 25-Ward model

In order to deeply understand the factors affecting the distribution of public wifi in Toronto, I will study the population and income of each ward in Toronto on this basis. Therefore, the regional profile dataset (Data 2021), based on the 2021 Census data, was also included in the analysis. The dataset, published by City Planning, was last updated on January 3, 2024. The dataset contains demographic, social and economic information such as income and population for each ward.[@toronto-ward-profiles]

Ward profile data is stored in an Excel file with multiple tabs. The relevant data used for this analysis is included as a variable in the first TAB, the 2021 Census. Therefore, only the data from this TAB is downloaded for analysis. Further data cleansing was done to transform the data, retaining only information related to income and population levels in each ward (see @tbl-ward). Based on the Toronto Open Data Portal, the package has a CSV file, 25 ward names and numbers, which contains the mapping between ward codes and ward names. I create statistic to visualize these data, using kable() from knitr to create @tbl-ward.

```{r}
#| label: tbl-ward
#| tbl-cap: Sample of cleaned Ward of Toronto data
#| echo: false

cleaned_ward_data %>%
  head(10) %>%
  kable(
    col.names = c("Population", "Income", "Ward Number"),
    digits = 1
  )

```

There are 25 wards in the City of Toronto. Based 2021 census data, the wards with the highest population are:
Etobicoke-Lakeshore (Ward 3) at 139920, Spadina-Fort York (Ward 10) at 135400, and Etobicoke Centre (Ward 2) at 117200. The wards with the lowest population are: Scarborough North (Ward
23) at 94025, Don Valley East (Ward 16) at 94335, and Don Valley West (Ward
15) at 101025.

The wards with the highest household income are: Spadina-Fort York (Ward 10) at $ 80730,
Toronto Centre (Ward 13) at $68965, and Etobicoke-Lakeshore (Ward 3) at $65575. The
wards with the lowest household income are: Humber River-Black Creek (Ward 7) at $65,458,
York South-Weston (Ward 5) at $67,964, and Scarborough Centre (Ward 21) at $70,624.

# Conclusion {#sec-conclusion}

## Demand

**Societal need for public Wi-Fi differs according to building types**

The type of building can often reflect the different needs of modern people for WiFi, and having more public WiFi coverage proves that it has a stronger degree of Internet digitization needs.

For example, libraries have embraced technology, expanding their offerings beyond books to include a rich array of digital resources. Public WiFi is now a vital tool for students, researchers, and casual readers, facilitating access to online databases, ebooks, and collaborative group projects. Moreover, Community Recreation Centre have transformed into dynamic digital hubs, serving as venues for social gatherings and work. Here, individuals can check emails, work remotely, and stay informed about current events, all while enjoying a cup of coffee.

## Population

**Higher population density correlates with increased distribution of free public WiFi**

```{r}
#| label: fig-1
#| fig-cap: Relationship between Population of each Ward and Amount of Free Public WiFi Location
#| echo: false
#| warning: false
#| message: false

analysis_data_1 <- read_csv(here::here("outputs/data/cleaned_ward_data.csv"))

analysis_data_1 |> 
  ggplot(aes(x = Population, y = WardNumber)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Population of each Ward",
       y = "Ward Number")

```

From this graph, we can clearly see that Etobicoke-Lakeshore (Ward 3) has the greatest population, and Spadina-Fort York (Ward 10) has the second greatest population.

In fact, this is similar to the distribution of free public WiFi in Toronto, but it does not exactly match. Ward 3, as the most populous area, has 13 public WiFi, which is a relatively large number in all Wards, and Ward 10 has 5, which is an insignificant number. But overall, for different Wards, a higher population slightly results in a higher amount of public WiFi.

## Income

**Income levels have limited influence on availability of free public Wi-Fi**

```{r}
#| label: fig-2
#| fig-cap: Relationship between Income of each Ward and Amount of Free Public WiFi Location
#| echo: false
#| warning: false
#| message: false

analysis_data_1 <- read_csv(here::here("outputs/data/cleaned_ward_data.csv"))

analysis_data_1 |> 
  ggplot(aes(x = Income, y = WardNumber)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Income of each Ward",
       y = "Ward Number")

```

From this graph, we can clearly see that Spadina-Fort York (Ward 10) has the greatest average income, Toronto Centre (Ward 13) has the second greatest average income, and Etobicoke-Lakeshore (Ward 3) has the third greatest average income.

We can conclude that there is no observable correlation between the number of Toronto public WiFi and the average household income in a ward.

@@

All in all, a significant number of people are increasingly relying on public Wi-Fi when they're out and about, underscoring their need for instant, hassle-free connectivity. Whether for work, social interaction or navigation, the modern lifestyle requires a seamless Internet experience, and the increasing number of academic and recreational projects joining the ranks of the application of public free WiFi is undoubtedly a strong support for the development of public WiFi in Toronto.

Furthermore, our data analysis indicates that while population size does exert a modest impact on the dissemination of free WiFi, the average income of a region lacks sufficient influence to be regarded as a determining factor in the allocation of public WiFi.
